Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-20 submission/manuscript.Rmd
edit discussion p1

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We observed global dysregulation of tissue-enriched genes in the _Kdm5c_-KO brain and EpiLCs, including substantial misexpression of testis, liver, muscle, and ovary-enriched genes. The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells, rather than somatic cells. We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis. We demonstrated female EpiLCs are more sensitive to germline gene de-repression, as the number germline genes misexpressed and their degree of dysregulation was greater in _Kdm5c_-KO females compared to males. KDM5C is enriched at germline gene promoters in EpiLCs but is bound to only a subset of germline genes transcribed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly expressed through indirect mechanisms. Finally, we found KDM5C is dynamically regulated during the differentiation of ESCs in EpiLCs and promotes long-term germline gene silencing by aiding the placement of DNA methylation at CpG islands in EpiLCs. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, we unveiled unique regulatory mechansisms governing distinct germline gene classes in somatic lineages. Altogether, these results provide molecular footholds that in the future can be exploited to test the contribution of ectopic germline genes upon neurodevelopment.

It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. Emerging evidence suggests KDM5C might negatively regulate totipotency genes, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023]. However, out of the four _Dazl_ regulators characterized, KDM5C was the only factor whose repression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023]. We found that while KDM5C also repressess _Dazl_ in EpiLCs, _Kdm5c_-KO EpiLCs did not express 2-cell specific genes. This suggests the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression. _Dazl_ and other "self-renewal" germline genes are typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021], but can become reactivated to form primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011]. Yet, _Kdm5c_-KO EpiLCs properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Proper _Otx2_ expression, coupled with normal EpiLC cellular morphology and _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.

>>>>

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity and characterized substantial misexpression of testis, liver, muscle, and ovary-enriched genes within the _Kdm5c_-KO brain. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expresses genes important for late spermatogenesis. We demonstrated that while _Kdm5c_ mutant sex did not influence the whether sperm or egg genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression. 
Germline-enriched mRNAs can become aberrantly expressed in _Kdm5c_-KO cells indirect of KDM5C, as KDM5C is enriched at germline gene promoters in EpiLCs but only bound to a subset of germline-enriched DEGs. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled unique repressive mechansisms governing distinct classes of germline gene in somatic lineages. Furthermore, these results provide molecular footholds that can be exploited to test the ultimate contribution of ectopic germline gene expression upon neurodevelopment.

It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. Emerging evidence suggests KDM5C might negatively regulate totipotency genes, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023]. However, out of the four _Dazl_ regulators characterized, KDM5C was the only factor whose repression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023]. We found that while KDM5C also repressess _Dazl_ in EpiLCs, _Kdm5c_-KO EpiLCs did not express 2-cell specific genes. This suggests the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression. _Dazl_ and other "self-renewal" germline genes are typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021], but can become reactivated to form primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011]. Yet, _Kdm5c_-KO EpiLCs properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Proper _Otx2_ expression, coupled with normal EpiLC cellular morphology and _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
